---
title: "ST558 Project1"
author: "Xinyu Hu"
date: "6/12/2020"
output: 
  rmarkdown::github_document:
    toc: yes
always_allow_html: true
---


# Before Everything
This is a project that goes through the pipeline from extracting data from public API, to processing in the RStudio, and to upload directly to the Github as README. Data was in JSON format when imported, which provides a great learning opportunity for JSON files; The output was directly push to the Github, which is also a great practice of what we learned in the class. It's an interesting and steep learning curve, and I love it quite a bit.

# Describe JSON
## What is Jason data?
Based on Wikipedia, [JSON (Jave Script Object Notation)](https://en.wikipedia.org/wiki/JSON) is an open standard lightweight file format for data interchange. It uses human-readable text to store and transmit data objects. It is a very common data format, with a diverse range of applications.Data that follows JSON format is JSON data.  

## Where does it get used?
According to [tutorialspoint](https://www.tutorialspoint.com/json/json_quick_guide.htm#:~:text=JSON%20format%20is%20used%20for,used%20with%20modern%20programming%20languages.), it is used while writing JavaScript based applications that includes browser extensions and websites. For example,transmit data between a server and web applications;Web services and APIs use JSON format to provide public data;serializing and transmitting structured data over network connection;used with modern programming languages.  

## Why it is a good way to store data?
Firstly, it is light-weighted. So before JSON, web services used XML as primary data format for transmitting, but after JSON appeard, it became the preferred formate. Secondly, JSON is easier for both humans and machines to understand, since its syntax is minimal and its structure is predictable. JSON was meant to carry structured information between programs from the very beginning. Plus, the nature of "human readable" also play a strong role that make JSON a good way to store data, which makes JSON not as confusing as XML. For more detail, feel free to check out [The Rise and Rise of JSON](https://twobithistory.org/2017/09/21/the-rise-and-rise-of-json.html) and [JSON.ORG](https://www.json.org/json-en.html).  

# Discuss JSON R-Packages
There are 3 packages in R that work with JSON file. They are [rjson](https://cran.r-project.org/web/packages/rjson/rjson.pdf), [jsonlite](https://cran.r-project.org/web/packages/jsonlite/jsonlite.pdf) and [RJSONIO](https://cran.r-project.org/web/packages/RJSONIO/RJSONIO.pdf). Also, there are package like [tidyjson](https://cran.r-project.org/web/packages/tidyjson/vignettes/introduction-to-tidyjson.html) that also can serve similar functionality.  `rjsonio` allows R objects to be inserted into Javascript/ECMAScript/ActionScript code and allows R programmers to read and convert JSON content to R objects.  `rjson`'s performance is now similar to this package, and perhaps slightly faster in some cases. `jsonlite` Started out as a fork of `rjsonio`, but has been completely rewritten in recent versions. The package offers flexible, robust, high performance tools for working with JSON in R and is particularly powerful for building pipelines and interacting with a web API.  

For this project, I choose `jsonlite` as it is the most familiar JSON package to me. Also, its fast speed is appealing.  

# Read-in JSON data
This is my step by step process to load in data from [NHL Franchise API](https://gitlab.com/dword4/nhlapi/-/blob/master/records-api.md).  
Step1: Load library    
```{r ,echo=TRUE,warning=FALSE,message=FALSE}
library("httr")
library("jsonlite")
```
Step2: Prepare Parent URL
```{r ,echo=TRUE,warning=FALSE,message=FALSE}
full_url <- "https://records.nhl.com/site/api"
```
Step3: Obtain Franchise data
```{r ,echo=TRUE,warning=FALSE,message=FALSE}
franchise <- fromJSON(paste(full_url,"/franchise",sep=""))
franchise <- franchise[[1]]
```
Step4: Obtain Franchise team totals
```{r ,echo=TRUE,warning=FALSE,message=FALSE}
franchise_team_totals <- fromJSON(paste(full_url,"/franchise-team-totals",sep=""))
franchise_team_totals <- franchise_team_totals[[1]]
```

Step5: Get the season record, goalie record and skater record for Franchise 1. Using this step, we can obtain the data columns from those 3 tables provided by the API.
```{r ,echo=TRUE,warning=FALSE,message=FALSE}
season_records_all <- fromJSON(paste(full_url,"/franchise-season-records?cayenneExp=franchiseId=1",sep=""))
season_records_all <- season_records_all[[1]]

goalie_records_all <- fromJSON(paste(full_url,"/franchise-goalie-records?cayenneExp=franchiseId=1",sep=""))
goalie_records_all <- goalie_records_all[[1]]

skater_records_all <- fromJSON(paste(full_url,"/franchise-skater-records?cayenneExp=franchiseId=1",sep=""))
skater_records_all <- skater_records_all[[1]]
```

Step6: loop through all the franchise (2 to 38), to obtain the season record, goalie record and skater record.  

```{r ,echo=TRUE,warning=FALSE,message=FALSE}
for (i in 2:38){
  # extract data from api
  season_records <- fromJSON(paste(full_url,"/franchise-season-records?cayenneExp=franchiseId=",i,sep=""))
  season_records <- season_records[[1]]
  
  goalie_records <- fromJSON(paste(full_url,"/franchise-goalie-records?cayenneExp=franchiseId=",i,sep=""))
  goalie_records <- goalie_records[[1]]
  
  skater_records <- fromJSON(paste(full_url,"/franchise-skater-records?cayenneExp=franchiseId=",i,sep=""))
  skater_records <- skater_records[[1]]
  
  # append to the giant table
  season_records_all <-rbind(season_records_all,season_records)
  goalie_records_all <-rbind(goalie_records_all,goalie_records)
  skater_records_all <-rbind(skater_records_all,skater_records)
  
  # remove temporary table after appending
  rm(season_records,goalie_records,skater_records) 
}
```

Step7: Check the data we read (first 5 columns, first 5 rows).
```{r ,echo=TRUE,warning=FALSE,message=FALSE}
head(franchise[,1:5],n=5)
head(franchise_team_totals[,1:5],n=5)
head(season_records_all[,1:5],n=5)
head(goalie_records_all[,1:5],n=5)
head(skater_records_all[,1:5],n=5)
```

# Exploratory Analysis

## Load Packages  
Lots of packages are used in the analysis section. `ggplot2` is a package with a spectrume of data visualization charts. It is very versatile and helpful. `dplyr`, `tidyr` and `tidyverse` are packages super helpful for data manipulation. And `DT` are used to create user-friendly data table in the output.  

```{r ,echo=FALSE,warning=FALSE,message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyverse)
library(DT)
library(tidyr)
library(knitr)
```

## Franchise Table Analysis
There is not so much information provided by the francise table. A very interesting point is to look at when is the first season for each franchise. In order to do so, I used `substr` function to get the first 4 digits, which is the year of first half. Then we plot the first year in a histogram.
```{r ,echo=TRUE,warning=FALSE,message=FALSE}
franchise$firstSeasonId_New <- as.numeric(substr(franchise$firstSeasonId,1,4))

FirstSeason_plot<-ggplot(data=franchise,aes(x=firstSeasonId_New))
FirstSeason_plot + geom_histogram(binwidth = 10, color="black", fill="blue", size=2)+
  labs(x = "Year of First Season")
```
I found that there are 2 time in the history that lots of franchises are created. One is between 1915 and 1925, during which 9 franchises were createed; Another time is between 1965 and 1975, totally 12 franchises were created then. After 1975, there remains a stream of Franchise creation, but not as many as those 2 peak time.  

Now I will create a table to identify franchise created before 1950, 1950 to 1980 and after 1980. Those 3 categories will be assigned as **Long Existing Franchise**,**Mid Existing Franchise**, and **Short Existing Franchise**  This table will be used later.
```{r ,echo=TRUE,warning=FALSE,message=FALSE}
FranchiseTenure <- franchise %>% select(id,firstSeasonId_New)
FranchiseTenure["FranchiseTenure"] <- ifelse(FranchiseTenure$firstSeasonId_New <= 1950, "Long Existing Franchise",
                             ifelse(FranchiseTenure$firstSeasonId_New <= 1980, "Mid Existing Franchise",
                             ifelse(FranchiseTenure$firstSeasonId_New > 1980, "Short Existing Franchise",
                                    "Missing Tenure")))
```

## Franchise Team Totals Analysis
This table offer a wide varitey of record about each franchise, by game type. In order to understand the winning rate for home and road games, several new variables are created. To be exact, Game Count and Win Rate are created at Overall, Home and Road levels.  

```{r ,echo=TRUE,warning=FALSE,message=FALSE}
franchise_team_totals['Overall_Game_Count'] <- rowSums(franchise_team_totals[,c("wins","losses","ties")],na.rm = T)
franchise_team_totals['Overall_Win_Rate'] <- round(franchise_team_totals$wins/franchise_team_totals$Overall_Game_Count,2)
franchise_team_totals['Home_Game_Count'] <- rowSums(franchise_team_totals[,c("homeWins","homeLosses","homeTies")],na.rm = T)
franchise_team_totals['Home_Win_Rate'] <- round(franchise_team_totals$homeWins/franchise_team_totals$Home_Game_Count,2)
franchise_team_totals['Road_Game_Count'] <- rowSums(franchise_team_totals[,c("roadWins","roadLosses","roadTies")],na.rm = T)
franchise_team_totals['Road_Win_Rate'] <- round(franchise_team_totals$roadWins/franchise_team_totals$Road_Game_Count,2)
```

With all those variables created, I create a table with only interesting metrics and sort them by franchise ID and game type ID.  
```{r ,echo=TRUE,warning=FALSE,message=FALSE}
franchise_team_SimpTotals <- franchise_team_totals %>% select(franchiseId,gameTypeId,Overall_Game_Count,Overall_Win_Rate,
                                                              Home_Game_Count,Home_Win_Rate,Road_Game_Count,Road_Win_Rate)%>% 
                                                    arrange(franchiseId,gameTypeId)
```  

Using the `datatable` function from `DT` package, I make a user-friendly datatable.
```{r ,echo=TRUE,warning=FALSE,message=FALSE}
datatable(franchise_team_SimpTotals,rownames = FALSE)
```  

This is a table that capture win rate for different senarios. Though not so much insight could be derived from this table, it serves as good source of data for further interests.   

## Season Record Analysis
Season record table provide lots of info related to seasonal performance of each franchise. The first step I would take is, to get some basic summary about win and loss streaks at home and at road. (I exclude *Montreal Wanderers* as its data contains numerous missing)  

```{r ,echo=TRUE,warning=FALSE,message=FALSE}
subtable<-round(apply(season_records_all[season_records_all$franchiseName!="Montreal Wanderers",c("homeWinStreak","homeLossStreak","roadWinStreak","roadLossStreak")],MARGIN=2,summary),2)
kable(subtable[ ,1:4])  
```   

Here I find that the average home loss streak (7.59) and road win streak (7.43) is very close. This makes sense because One franchise's home loss will be the counterpart's raod win. The similarity also exist when compare average home win streak and road loss streak.  

One step further, I am curious teams whose road loss streak less than 12 will have different home win streak distribution. So first, I created a character variable named `roadLossStreak_Category`, assigning values depending one if the road loss streak is less than 12. Then create a side-by-side histogram to compare the home win streak.  

```{r ,echo=TRUE,warning=FALSE,message=FALSE}
season_records_all["roadLossStreak_Category"] <- ifelse(season_records_all$roadLossStreak <12, "Less than 12","No Less than 12")

WinLossStreak_plot<-ggplot(data=season_records_all,aes(x=homeWinStreak))
WinLossStreak_plot + geom_histogram(binwidth = 1,color="black", fill="blue", size=2)+
                     labs(x = "Home Win Streak")+
  facet_grid(cols=vars(roadLossStreak_Category))
```  

With limited number of franchise, both distributions look robust. But based on this table, I think the home win streak of franchiases with Less than 12 road loss streak (left histogram) are centering aroud 9, while the "No Less than 12" group (right histogram) are centered at 11. Also, the left histogram looks more bell-shaped, while the right one's peak are right at 5 and skewed to the right.  

## Goalie Analysis  
Goalie table from the API returns information about each goalie. It includes both active and non-active goalies. Basically, I am interested in how many active and non active player by franchise tenure. So I created a contingency table using the code below.*(firstly, I reassign value of activePlayer variable to be more reader friendly)*    

```{r ,echo=TRUE,warning=FALSE,message=FALSE}
goalie_records_all <- left_join(goalie_records_all,FranchiseTenure,by = c("franchiseId" = "id"))
goalie_records_all$activePlayer <- ifelse(goalie_records_all$activePlayer=="TRUE",'Active Goalie','Non-Active Goalie')
goalie_records_all%>%
  group_by(FranchiseTenure, activePlayer)%>%
  summarise(n=n())%>%
  spread(activePlayer, n)%>%
  kable()
```  

I am very interested to learn goalie's most win by franchise and active status. So I created a side-by-side box plot as below. 
```{r ,echo=TRUE,warning=FALSE,message=FALSE}
goalie_records_all$franchiseId <- as.factor(goalie_records_all$franchiseId)
goalie_records_plot<-ggplot(goalie_records_all,aes(x=franchiseId,y=mostWinsOneSeason))
goalie_records_plot + geom_boxplot()+
  facet_grid(rows =vars(goalie_records_all$activePlayer))+
  labs(title="Box-Plot: Seasonal Most Win By Goalie Type")
```  

Comparing active goalie vs non-active goalie, there isn't so much insight to obtain visually from this plot. The only trend I can observe is, the active goalie are higher in *most win one season* in terms of averages than non-active goalie in most teams. When we focus on franchise vs franchise, I do see a difference in distribution of *most win one season*. While most are right tailed, some franchise have average *most win one season* being around 5 (franchise 25 for example), and others being around 15 (franchise 12 for example).  

# Skater Record Analysis
Skater Record table is a table that provides abundant info about skaters. Just like Goalie, I'm interested in how many active and non-active each of the franchise has right now. So I create a contigency table to do the count. *(firstly, I reassign value of activePlayer variable to be more reader friendly)* 
```{r ,echo=TRUE,warning=FALSE,message=FALSE}
skater_records_all <- left_join(skater_records_all,FranchiseTenure,by = c("franchiseId" = "id"))

skater_records_all$activePlayer <- ifelse(skater_records_all$activePlayer=="TRUE",'Active Skater','Non-Active Skater')
skater_records_all$activePlayer <- as.factor(skater_records_all$activePlayer)
skater_records_all$mostPointsOneGame <- as.numeric(skater_records_all$mostPointsOneGame)
skater_records_all%>%group_by(FranchiseTenure, activePlayer)%>%
                    summarise(n=n())%>%
          spread(activePlayer, n)%>%
                    kable()
```

Lastly, I am interesed in the relationship between Most Assists One Season and Most Goal One Season for each of the skater. Especially, I want to know how *most point one game* can influence on the relationship. So I make a scatter plot, with color to indicate *most point one game*. The color theme comes from `wesanderson` package.
```{r ,echo=TRUE,warning=FALSE,message=FALSE}
library(wesanderson)
skater_records_plot<- ggplot(skater_records_all,aes(x=mostAssistsOneSeason, y=mostGoalsOneSeason, color=mostPointsOneGame)) 
skater_records_plot + geom_point()+scale_color_gradientn(colours = rainbow(8))

```  

There are so many observation can be made by this plot. First, we do see that as *most points one game* increase from orange to purple, the data points are moving from the bottom left to the top right. This suggests a positive correlation among *most points one game*, *most goal one season* and *most assist one season*. This finding verifies that the best skater can be doing well in all aspects. Second finding is, as *most points one game* increases, the data points are scatter in a wider zone. For example, for those orange dots, it all nested ina 5 by 5 zone, while for green dots, the zone is from 5 to 50 horizontally, and 5 to 45 vertically. Thirdly, the slop looks very different as *most points one game* increases. The slope of purple dots are more flat, while the slope for green dots are more steep.


# In the End
If you have any comments, questions or suggestions, please email me at [xhu23@ncsu.edu](mailto:xhu23@ncsu.edu). I'm sure there are so many interesting stories that I havn;t touched yet!
